Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm

### 2014-08-20 paint-contour-tree.tex
small changes

Let $\mathcal{L}$ be a light path tree of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
Our goal is to use \Lem{geometric}  to show that $\sum_{v\in \mathcal{L}} |H_v| = O(|H_r|)$, namely the total sum of the sizes of the heaps 
at the roots of all paths in the light path tree is on the order of the size of the heap at the root of $r$.  
ADD THE ARGUMENT saying how this fact will be used.

@@

Let $W= \sum_{v\in \mathcal{L}} w(v) \geq \sum_{v\in \mathcal{L}} |H_v|$.  Suppose we could show that conversely $W=O(H_r)$.  
Then we would have $\sum_{v\in \mathcal{L}} H_v = O(H_r)$, which was our original goal.  So proving $W=O(H_r)$ is now our new goal.

To this end, first observe that $|H_r|\geq w(r)-\sum_{v\in \mathcal{L}} |v|$.  This holds by the same argument as in \Lem{adjacent}, namely, 
$w(r)$ by definition counts all the contributing heavy path heaps and initial leaf heaps that contribute to the heap at $r$ and $\sum_{v\in \mathcal{L}} |v|$
is an upper bound on the number of vertices that could have been removed from one of these heaps before reach the root of $r$ (since a vertex in a heap can only 
get removed if it appears on one of the paths).

@@

\begin{claim}
$|H_r|\geq w(r)-\sum_{v\in \mathcal{L}} \sqrt{H_v}/100 \geq w(r)-\sum_{v\in \mathcal{L}} \sqrt{w(v)}/100$
\end{claim}

We will need the following lemma.

\begin{lemma}
\label{lem:distribution}
Let $C\subseteq \mathcal{P}$ be the subset of paths that are leaves in $\mathcal{L}$.

Then $\sum_{v\in \mathcal{P}} \sqrt{w(v)}/100 \leq \sum_{c\in C} \sqrt{w(c)}/20$
\end{lemma}

\begin{proof}
The idea is to recursively redistribute the contribution to the sum of each internal vertex to its children, until all the weight sits on the children.
Specifically, let $v$ be some path in $\mathcal{L}$.  Its contribution to the sum is $\sqrt{w(v)}/100$, which we will call the value at $v$.  
We redistribute this value to its children, 
in proportion by the weight of each child.  For example, if $v$ has a child $x$ then the fraction it gets assigned is $w(x)/w(v)$ 
(recall that by definition $w(v)$ is the sum of the weights of its children and so the total value will get reassigned to the children).

Suppose we keep redistributing the value of $v$ until it reaches the leaves of $\mathcal{L}$.  Let's see how much gets assigned to a given leaf.
Specifically, let $u_0, u_1, \ldots, u_k$ be a path in $\mathcal{L}$ (of paths from the set $\mathcal{P}$), such that $v=u_0$,
$u_{i+1}$ is child of $u_i$ for all $i$, and $u_k$ is a leaf of $\mathcal{L}$.

@@

Then by the above calculation and \Lem{geometric}, the total reassigned to $u_m$ is

\Ben{There is an issue here.  \Lem{geometric} is in terms of the heaps not the weights.}

@@

The claim now follows by summing this quantity over the children of $\mathcal{L}$.
\end{proof}

By definition, $w(r) = \sum_{c\in C} w(c)$, where $C$ is the subset of leaves of $\mathcal{L}$.  
Therefore combining the above lemma and claim gives the following.

\begin{claim}
 $|H_r|\geq w(r)/2$
\end{claim}

\begin{lemma}
 $W=O(H_r)$
\end{lemma}
\begin{proof}
MISSING.  Is this what we are claiming?  I don't see the proof.
\end{proof}
%
%
% So $w(r)$ upper bound $H_r$, but by how much?  Consider a path $v\in \mathcal{L}$.  
% As we move from the leaf to the root of $v$ we now want to understand 
% how the heap can decreases.  Now the only way a vertex in the heap can leave that 
% heap is if that vertex is one of the vertices we traverse in $v$, i.e. the total 
% loss in vertices from the heap along $v$ is bounded by $|v|$.  
% Therefore, $|H_r|\geq W-\sum_{v\in \mathcal{L}} |v|$.  
% (This is an extension of \Lem{adjacent}, and probably should be merged with it.)
% As $\mathcal{L}$ consists of only light paths we have $|v|\leq \sqrt{|H_v|}/100 \leq \sqrt{w(v)}/100$, 
% and therefore $|H_r|\geq W-\sum_{v\in \mathcal{L}} \sqrt{w(v)}/100$.

>>>>

Let $\mathcal{L}$ be a light path tree of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
We now define a weight function $w(v)$ for all $v\in \mathcal{L}$.  Such that $w(v)=\Theta(|H_v|)$.  
For this alternative quantity we will be able to show BLAH BLAH BLAH.

%Our goal is to use \Lem{geometric}  to show that $\sum_{v\in \mathcal{L}} |H_v| = O(|H_r|)$, namely 
%the total sum of the sizes of the heaps at the roots of all paths in the light path tree is on the 
%order of the size of the heap at the root of $r$.  ADD THE ARGUMENT saying how this fact will be used.

@@

%Let $W= \sum_{v\in \mathcal{L}} w(v) \geq \sum_{v\in \mathcal{L}} |H_v|$.  
%Suppose we could show that conversely $W=O(H_r)$.  Then we would have 
%$\sum_{v\in \mathcal{L}} H_v = O(H_r)$, which was our original goal.  
%So proving $W=O(H_r)$ is now our new goal.

We now would like to show conversely that $w(v)= O(|H_v|)$.  
To do so we first introduce some notation.  
Let $C\subseteq \mathcal{P}$ be the subset of paths that are leaves in $\mathcal{L}$.
For any path $v$, let $\mathcal{L}_v$ (resp. $C_v$) be the subset of paths in $\mathcal{L}$ (resp $C$) 
that are contained in the subtree of $\mathcal{L}$ which is rooted at $v$.

First, observe that $|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} |u|$.  
This holds by the same argument as in \Lem{adjacent}, namely, $w(u)$ by definition 
counts all the contributing heavy path heaps and initial leaf heaps that contribute 
to the heap at $v$ and $\sum_{u\in \mathcal{L}_v} |u|$ is an upper bound on the number 
of vertices that could have been removed from one of these heaps before reaching the 
root of $v$ (since a vertex in a heap can only get removed if it appears on one of 
the paths).

@@

\begin{claim}
$|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{|H_u|}/100 \geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100$
\end{claim}

By the above claim, if we can prove $\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \leq w(v)/c$ for some constant $c>1$, 
then we will have proven that $|H_v| = \Theta(w(v))$.

Recall that \Lem{geometric} told us that that $|H_{u}|\leq (3/5) |H_{u'}|$, if $u'$ is the parent $u$ in $\mathcal{L}$. 
The following lemma shows that if we had a similar statement in terms of the weights rather than the heap sizes, 
then we get the desired bound on the sum.

\begin{lemma}
\label{lem:distribution}
Let $v$ be a path in $\mathcal{L}$.  Suppose that for any two paths $x,y\in \mathcal{L}_v$ 
such that $y$ is the parent of $x$, $w(x)\leq (4/5) w(y)$.
Then $\sum_{u\in \mathcal{P}_v} \sqrt{w(u)}/100 \leq \sum_{c\in C_v} \sqrt{w(c)}/20$
\end{lemma}

\begin{proof}
The idea is to recursively redistribute the contribution to the sum of each internal vertex to its children, 
until all the weight sits on the leaves.  Specifically, let $v$ be some path in $\mathcal{L}$.  Its 
contribution to the sum is $\sqrt{w(v)}/100$, which we will call the value at $v$.  We redistribute this 
value to its children, in proportion by the weight of each child.  For example, if $v$ has a child $x$ then 
the fraction it gets assigned is $w(x)/w(v)$ (recall that by definition $w(v)$ is the sum of the weights of 
its children and so the total value will get reassigned to the children).

Suppose we keep redistributing the value of $v$ until it reaches the leaves of its subtree in $\mathcal{L}$.  
Let's see how much gets assigned to a given leaf.  Specifically, let $u_0, u_1, \ldots, u_k$ 
be a path in $\mathcal{L}$ (of paths from the set $\mathcal{P}_v$), such that $v=u_0$, $u_{i+1}$ 
is child of $u_i$ for all $i$, and $u_k$ is a leaf of $\mathcal{L}$.

@@

By assumption from the lemma statement we have $w(u_{i+1})\leq (3/5) w(u_i)$.  Therefore by the above calculation, 
the total reassigned to $u_m$ is

@@

The claim then follows by summing this quantity over the children of $\mathcal{L}$.
\end{proof}

By combining the above lemma and claim we have the following.

\begin{claim}
For any path $v\in \mathcal{L}$ if it holds that 
$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, then

\[
|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \geq w(v)- \sum_{c\in C_v} \sqrt{w(c)}/20 \geq \frac{4}{5} w(v)
\]
\end{claim}

\begin{lemma}
Let $\mathcal{L}$ be the light path tree of some path $r$.  
Then for any $v\in \mathcal{L}$, we have $|H_v|\geq \frac{4}{5} w(v)$.
\end{lemma}
\begin{proof}
We will actually prove the stronger statement that for any $v\in \mathcal{L}$, 
we have $|H_v|\geq \frac{4}{5} w(v)$ and $w(v)\leq (4/5) w(p)$ for any two paths 
$x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$.

Suppose otherwise, and let $v$ be the lowest path in $\mathcal{L}$ such that this fails 
(i.e. the statement holds for all descendants of $v$ in $\mathcal{L}$).  Then for any 
two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, 
we have that $w(x)\leq (4/5) w(y)$ and so by \Lem{geometric},

%
%
% So $w(r)$ upper bound $H_r$, but by how much?  Consider a path $v\in \mathcal{L}$.  
% As we move from the leaf to the root of $v$ we now want to understand 
% how the heap can decreases.  Now the only way a vertex in the heap can leave that 
% heap is if that vertex is one of the vertices we traverse in $v$, i.e. the total 
% loss in vertices from the heap along $v$ is bounded by $|v|$.  
% Therefore, $|H_r|\geq W-\sum_{v\in \mathcal{L}} |v|$.  
% (This is an extension of \Lem{adjacent}, and probably should be merged with it.)
% As $\mathcal{L}$ consists of only light paths we have $|v|\leq \sqrt{|H_v|}/100 \leq \sqrt{w(v)}/100$, 
% and therefore $|H_r|\geq W-\sum_{v\in \mathcal{L}} \sqrt{w(v)}/100$.
